---
title: "DEI and Party Identification: Who is more likely to think DEI practices are positive?"
author: "Zoey Jiang"
description: "A look into how political party affiliation and attitudes toward Diversity, Equity, and Inclusion initiatives in the workplaces may differ."
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

This study examines the connection between political party affiliation and opinions toward workplace diversity, equity, and inclusion (DEI) programs. The public's perception of formal DEI programs has become a central topic of political and cultural discussion as organizations adopt them more frequently. To comprehend how business policies interact with broader social identities in the United States, it is crucial to determine whether these sentiments are political.The main research question is: **Does a person's political party identification have a substantial impact on whether they think DEI practices in the workplace are a positive or negative development for the workforce?** 

My primary hypothesis is that, in comparison to respondents who identify as or lean Republican, those who identify as Democrats or lean Democratic will be substantially more likely to report that DEI practices have a positive impact on the workplace. This field of study is crucial because it shows how partisan perspectives can influence how people view workplaces and equity-based organizational reforms.

```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(haven)
library(modelsummary)
library(knitr)

dataset <- read_sav("ATP W154.sav")
head(dataset)

clean <- data.frame(
  F_PARTYLN_FINAL = sample(c("Republican", "Democrat"), 1000, replace = T), 
  DEIAFFECT = sample(c(1,2,3,4,98,99), 1000, replace = T)
)

# cleaning more

workingclean <- clean |> 
  filter(DEIAFFECT <= 4) |> 
  mutate(dei_score = 5 - DEIAFFECT) |> 
  mutate(party = factor(F_PARTYLN_FINAL, levels = c("Republican", "Democrat"))) |> 
  select(party, dei_score)

head(workingclean)
```

## Distribution of Party Identification from Sample

By taking a sample from the Pew Research Center's 

```{r, echo=FALSE}
ggplot(workingclean, aes(x = party, fill = party)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Distribution of Party Identification from Sample",
    x = "Political Party",
    y = "Number of Respondents",
    fill = "Political Party"
  ) + 
  scale_fill_manual(values = c("deepskyblue", "pink"))
```

This plot shows the overall distribution of party identification in the sample. It provides context for understanding how respondents are divided, or not, between the two major political parties.

## Distribution of DEI Attitudes

This analysis looks at the connection between political party affiliation and gender using data from the Pew datset 154. I specifically want to know if men are more likely than women to identify as Republicans instead of Democrats. I use a summary table to compare trends between genders after first visualizing the distribution of party identification among male respondents in order to investigate this.


```{r, echo=FALSE}
dei_dist <- ggplot(workingclean, aes(x = as.factor(dei_score))) +
  geom_bar(fill = "mediumpurple1") +
  labs(
    title = "Distribution of DEI Attitudes",
    x = "Favorability Score (1 = Bad, 4 = Good)",
    y = "Number of Respondents" 
  )
dei_dist
```
## Comparison by Party

```{r}
comp_party <- ggplot(workingclean, aes(x = party, y = dei_score, fill = party)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.1) +
  scale_fill_manual(values = c("Republican" = "pink","Democrat" = "deepskyblue")) +
  labs(
    title = "DEI Favorability by Political Party",
    x = "Party Affiliation",
    y = "DEI Favorability Score"
  )
comp_party
```

# Table Visualization

```{r, echo=FALSE}
tabledata <- workingclean |> 
  group_by(party) |> 
  summarize(
    `# of Respondents` = n(),
    `Mean Score` = round(mean(dei_score, na.rm = T), 2),
    `SD` = round(sd(dei_score, na.rm = T), 2),
    `Median` = median(dei_score, na.rm = T)
  )

table <- tabledata |> 
  knitr::kable(caption = "Summary of DEI Favorability Scores by Political Affiliation",
               align = "lcccc", 
               booktabs = T)
table
```

text about table

## Regression Model

text

```{r, echo=FALSE}
dei_reg <- lm(dei_score ~ party, data = workingclean)

modelsummary::modelsummary(dei_reg, 
                           title = "Linear Regression: Effect of Partisanship on DEI Attitudes",
                           gof_map = c("nobs", "r.squared"))
```

## Conclusion



## Code (R Studio)

```{r print-code, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

```